---
title: "BIOL 4800/6220_Fall2022Tutorial"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

# R basic

## Introduction

### R as a calculator
```{r}
1+200
```
```{r}
2^2
```
```{r}
7/2
```
```{r}
3487525*4097520
```
```{r}
cos(1)
```
```{r}
exp(0)
```
```{r}
1 == 1 #asking a question, is 1=1?
```
```{r}
1>= 1
```
```{r}
1>1
```

```{r}
1!= 2 #is 1 not equal to 2?
```

```{r}
x <- 10
x
```
```{r}
y <- x+x
y
```
### Vectorization (list-making)

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
log10(1:5)
```

### Environment 
```{r}
ls() #list of everything open/ created/ defined
```
```{r}
rm(x) #removes a variable, useful for clearing up memory space
```

```{r}

```
## Package Management 

```{r}
installed.packages()
```
```{r}
#install a package
#install.packages("Rtools", dependencies=TRUE)
```

## Project management
 1. set up management structure
 2. Discuss avoiding redundancy of files
 3.Make sure to use version control
```


```## Getting Help
?plot

```{r}
help(plot)

```
```{r}
vignette("FAQ-vegan")
```
```{r}
citation("vegan")
```

## Data and their formats






```{r}
cats <- data.frame(coat = c ("callico","tabby","black"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```
```{r}
write.csv(cats,file = "./cats.csv")
```
```{r}
cats
```
```{r}
cats$coat
```

```{r}
paste("my cat is", cats$coat, ", and it weighs" , cats$weight, " kg.", sep=" ")
```

### DATA TYPES

```{r}
typeof(cats$coat)
```

```{r}
typeof(cats$weight)
```
```{r}
typeof(cats$likes_string)

```
```{r}
typeof(1i)
```

```{r}
typeof(FALSE)
```
```{r}
typeof(TRUE)

```
```{r}
typeof(T)
```
```{r}
typeof(F)
```
```{r}
typeof(cats)
```

```{r}
file.show("./cats.csv")
```

```{r}
cats$likes_string <- as.logical(cats$likes_string)
```

```{r}
typeof(cats$likes_string)
```
```{r}
cats$likes_string
```

```{r}
z <- seq(10)
```

```{r}
class(z)
```

```{r}
seq(1,100, by=5)
```
```{r}
cats
```
```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby","tortoiseshell","black","tabby")
coats
```
```{r}
factor(coats)
```

```{r}
categories <- factor(coats)
class(coats)
class(categories)
```
```{r}
str(categories)
```

### Lists
```{r}
list_example <- list("numbers", numbers=1:10, data=TRUE)
list_example
```
```{r}
another_list<-list(1,"a",TRUE,1+1i)
another_list
```
```{r}
typeof(cats)
```
```{r}
class(cats)
```

```{r}
data.frame(list_example)
```


```{r}
cats
```
```{r}
cats[,3]
```

```{r}
cats[3,]
```

```{r}
cats[2:3,c(1,3)]
```


### Matricies 
```{r}
#a matrix is made for math
matrix_example <-matrix(0,ncol=5,nrow=3)
```
```{r}
matrix_example
```

```{r}
typeof(matrix_example)
```

```{r}
str(matrix_example) #turns into string
```

```{r}
dim(matrix_example)
```


```{r}
class(data.frame(matrix_example)) 
#save as data frame:
df_example <-data.frame(matrix_example)
df_example
```


###Subsetting
```{r}
p <- c(2.3,6.9,4.0,23,12)
p
```

```{r}
names(p) <- c('a','b','c','d','e')
p
```


```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
p[-3] #everything but 3
```

```{r}
p[c(-1,-5)]
```

```{r}
p[-(2:4)]
```

```{r}
p[c('a','c')]
```

```{r}
p[c(T,F,T,F,T)]
```

```{r}
#search
p[names(p) != 'c']
```

### Factors
```{r}
f <- factor(c('a','b','c','d','e'))
f
```

```{r}
#search
f[f =='a']
```

```{r}
f[1:3]
```

```{r}
f[f %in% c('b','c')]
#% is a special operator
```

```{r}
f[-3]
```

```{r}
f2 <- factor(c('a','a','d','c','c'))
f2
```

```{r}
f2[f2=='a']
```

```{r}
f2[f2 %in% c('a','c')]
```

```{r}
#git pull
```
### Matrices Cont.
```{r}
set.seed(65)
m<-matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```

```{r}
m[c(93,4)]
```
### Lists
```{r}
#xlist <-(a="BIOL6220" , b= seq(1 , 10 , by=0.5) , data="grade")
#data
```
```{r}
gp<- read.csv('~/BIOL6220_Fall22/practcomp2022/data/population_total.csv')
```
```{r}
head(gp, n=10L)
```

```{r}
head(gp[3], n=10L)

```

```{r}
nrow(gp)
```
```{r}
gp$X1799
```

```{r}
gp[1:3,2:5]
```
```{r}
gp[c(1,3), 2:5]
```

```{r}
gp[which(gp$X2059 <= 2 & gp$X2019 >=2)]
```
### Conditionals and flow

```{r}
n <-10
if (n<10) {
  print("n is <10")
} else if (n>10) {
  print("n is >10")
}else {
  print("I love moth man")
}
```
```{r}
for (n in seq(0,20)){
  n <-10
  if (n<10) {
  print("n is <10")
  } else if (n>10) {
  print("n is >10")
  }else {
  print("I love moth man")
  }
}
```
```{r}
g<-0
while(g<=10){
  print(paste(g,"is less than or equal to 10"))
  g <- g+1
}
```

```{r}

```
## Plotting and figures

```{r}
#install.packages(("ggplot2"))
```

```{r}
library(ggplot2)
```

```{r}
```
```{r}
ggplot(data=gp, mapping=aes(x=X2000, y=X2019)) +geom_point()
```

```{r}
library(ggplot2)
gapminder_all <- read_csv("data/gapminder_all.csv")
View(gapminder_all)
```

```{r}
ggplot(data=gapminder_all,mapping=aes(x=gdpPercap_1952))+geom_point()
```

```{r}
gapminder<-read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_point()

```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line(mapping=aes(color=continent))+geom_point()

```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.5)+scale_x_log10()
#log transformed data. now, we could run a linear model
#alpha tells us how densely packed the points are by changing the shading

```

```{r}
#linear model
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size=0.9, shape=9) +scale_x_log10()+geom_smooth(method = lm, color="gold")

```

```{r}
africas<-gapminder[gapminder$continent=="Africa",]
head(africas)
#subselecting data
```
```{r}
ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="blue") +
  facet_wrap(~country) +
  theme(axis.text.x=element_text(angle=45)) +
#changes the x axis so you can read the titles better
  labs(
    x="year",
    y="Life Expectancy",
    title="Life expectancy over time in African countries"
  )
# you can write code line by line as above for easier reading 
```

```{r}
AfricanLifeExp<-ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="blue") +
  facet_wrap(~country) +
  theme(axis.text.x=element_text(angle=45)) +
  labs(
    x="year",
    y="Life Expectancy",
    title="Life expectancy over time in African countries"
    )
```
```{r}
AfricanLifeExp
```
```{r}
ggsave(filename="AfricanLifeExp.png", plot=AfricanLifeExp, width=24, height = 40, dpi=300, units="cm")
```

```{r}
pdf(file= "results/AfricanLifeExp.pdf", width=24, height = 40)
plot(AfricanLifeExp)
dev.off()
#Turns off the creation of new files after its done
```

```{r}
write.table(gapminder, file="data/gapminder_web.csv", sep=",")
#save as csv. sep= choose what to separate by (comma, space, etc)
```

```{r}
write.csv(africas, file="data/gapminder_web_africas.csv")

```

